---
title: "Final Project - Metropolis-Hastings"
author: "Boxin Zhao"
date: "10/27/2018"
output: pdf_document
---

# 1. Algorithm description

My algorithm for M-H sampling is as below:

(1) Generate $x_{1}\sim{Uniform[0, 1]}$

(2) For t = 1 to (n.iteration - 1):
    (a) Generate $x^{*}\sim{Beta(x|cx_{t}, c(1-x_{t}))}$
    (b) Generate $u\sim{Uniform[0, 1]}$
    (c) Let $\alpha(x_{t},x^{*})=min\{\frac{Beta(x^{*}|6,4)Beta(x_{t}|cx^{*},c(1-x^{*}))}{Beta(x_{t}|6,4)Beta(x^{*}|cx_{t},c(1-x_{t}))}, 1\}$
        if $u<\alpha(x_{t},x^{*})$:
            $x_{t+1}=x^{*}$
        else:
            $x_{t+1}=x_{t}$

Below is the implementation:

```{r}
c <- 1
n.iteration <- 10000

sample.record <- numeric(n.iteration)

sample.record[1] <- runif(1)
for (t in 1:(n.iteration - 1)){
  xt <- sample.record[t]
  x.prop <- rbeta(1, c * xt, c * (1 - xt))
  u <- runif(1)
  alpha <- min(pbeta(x.prop, 6, 4) * pbeta(xt, c * x.prop, c * (1- x.prop)) / 
                 (pbeta(xt, 6, 4) * pbeta(x.prop, c * xt, c * (1- xt))), 1)
  if (u < alpha){
    sample.record[t + 1] <- x.prop
  } else {
    sample.record[t + 1] <- xt
  }
}
```

# 2. Evaluate the performance of the sampler. 

Let's now analysis the result.

First, a trace plot of this sampler and an autocorrelation plot, as well as a histogram of the draws.

```{r}
par(mfrow=c(1,3))  #1 row, 3 columns
plot(sample.record); acf(sample.record); hist(sample.record)  #plot commands
```

Compare the histogram of the draws with the target distribution

```{r}
tar.sample <- rbeta(n.iteration, 6, 4)

par(mfrow=c(1,2))
hist(sample.record); hist(tar.sample)
```

And the Kolmogorov¨CSmirnov test

```{r}
ks.test(sample.record, pbeta, 6, 4)
```

# 3. Re-run this sampler with c = 0.1, c = 2.5 and c = 10.

When c = 0.1

```{r}
c <- 0.1
n.iteration <- 10000

sample.record <- numeric(n.iteration)

sample.record[1] <- runif(1)
for (t in 1:(n.iteration - 1)){
  xt <- sample.record[t]
  x.prop <- rbeta(1, c * xt, c * (1 - xt))
  u <- runif(1)
  alpha <- min(pbeta(x.prop, 6, 4) * pbeta(xt, c * x.prop, c * (1- x.prop)) / 
                 (pbeta(xt, 6, 4) * pbeta(x.prop, c * xt, c * (1- xt))), 1)
  if (u < alpha){
    sample.record[t + 1] <- x.prop
  } else {
    sample.record[t + 1] <- xt
  }
}

par(mfrow=c(1,3))  #1 row, 3 columns
plot(sample.record); acf(sample.record); hist(sample.record)  #plot commands
```

When c = 2.5

```{r}
c <- 2.5
n.iteration <- 10000

sample.record <- numeric(n.iteration)

sample.record[1] <- runif(1)
for (t in 1:(n.iteration - 1)){
  xt <- sample.record[t]
  x.prop <- rbeta(1, c * xt, c * (1 - xt))
  u <- runif(1)
  alpha <- min(pbeta(x.prop, 6, 4) * pbeta(xt, c * x.prop, c * (1- x.prop)) / 
                 (pbeta(xt, 6, 4) * pbeta(x.prop, c * xt, c * (1- xt))), 1)
  if (u < alpha){
    sample.record[t + 1] <- x.prop
  } else {
    sample.record[t + 1] <- xt
  }
}

par(mfrow=c(1,3))  #1 row, 3 columns
plot(sample.record); acf(sample.record); hist(sample.record)  #plot commands
```

When c = 10

```{r}
c <- 10
n.iteration <- 10000

sample.record <- numeric(n.iteration)

sample.record[1] <- runif(1)
for (t in 1:(n.iteration - 1)){
  xt <- sample.record[t]
  x.prop <- rbeta(1, c * xt, c * (1 - xt))
  u <- runif(1)
  alpha <- min(pbeta(x.prop, 6, 4) * pbeta(xt, c * x.prop, c * (1- x.prop)) / 
                 (pbeta(xt, 6, 4) * pbeta(x.prop, c * xt, c * (1- xt))), 1)
  if (u < alpha){
    sample.record[t + 1] <- x.prop
  } else {
    sample.record[t + 1] <- xt
  }
}

par(mfrow=c(1,3))  #1 row, 3 columns
plot(sample.record); acf(sample.record); hist(sample.record)  #plot commands
```

I think when c = 10, the sampler is most effective. Because the shape of the histogram is most similar to the hist of target distribution, which indicates we need to discard less samples (or a early burn-in). If we analyze the properties of target distribution, we can draw the same conclusion. Because for a x drawn from $Beta(6, 4)$, the shape of $Beta(cx, c(1-x))$ is most similar to $Beta(6, 4)$ when c = 10 (among c = 0.1, 1, 2.5 or 10). In this way, we will have a higher acceptance probability and more efficient drawing process.

